Article Title

### 2014-05-15 plos.tex

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations, and other mechanisms \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than an equivalent number of separately recorded pairs of cells. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by only a few eigenvectors that affect all correlation coefficients. On the other hand, a correlation matrix can be specified using the \emph{partial correlations} between pairs of the recorded neurons. The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells.  When a large fraction of the interacting variables are observed, partial correlations more closely reflect the direct causal effects between components of complex systems than correlations. Partial correlations have been used to describe interactions between genes in functional genomics \cite{Schafer:2005, Peng:2009} and between brain regions in imaging studies \cite{Varoquaux:2012, Ryali:2012}. These opportunities have not yet been explored in neurophysiological studies, as most studies have only considered the distributions of pairwise correlations \cite{Zohary:1994, Bair:2001, Smith:2008, Ecker:2010}.

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.  The usual way to estimate the covariance matrix is to calculate the \emph{sample covariance matrix} $C_{\sf sample}$ from the empirical sample of observations $x(t),\; t=1,\ldots,n$ as
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ is the number of degrees of freedom in one neuron's activity within the sample ($\nu=n-1$ if observations are independent). For noise correlations, the true mean $\mu$ and the sample mean $\bar x$ are conditioned on the stimulus. In this study, we estimate the true mean, $\mu$, using the sample mean, $\bar x$, but seek a better estimate of the covariance matrix than the sample correlation matrix, $C_{\sf sample}$.

Given an arbitrary covariance matrix estimate $C$, the corresponding correlation matrix $R$ is calculated by normalizing $C$ by its diagonal (variance estimates):
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix. 
Similarly, the matrix of partial correlations $P$ is computed by normalizing the \emph{precision matrix} $C^{-1}$:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

To see that this is the case, partition the covariance matrix 
\begin{equation}
C = \begin{pmatrix}
C_{11} & C_{12} \\
C_{12}^\T & C_{22}
\end{pmatrix}
\end{equation}
so that $C_{11}$ is the $2\times 2$ covariance matrix of two cells of interest. Such partitioning can be done for any pair of cells by changing their order. 
Then the covariance matrix of the conditional distribution of these two cells is \cite{Anderson:2003}:
\begin{equation}
\hat C_{11} = C_{11}-C_{12}C^{-1}C_{12}^\T 
\end{equation}
The inverse of the conditional covariance matrix can be found in the corresponding partition of the precision matrix $C^{-1}$:
\begin{equation}
C^{-1}=\begin{pmatrix}
\hat C_{11}^{-1} &   -\hat C_{11}^{-1}C_{12}C_{22}^{-1} \\
-C_{22}^{-1}C_{12}^\T\hat C_{11}^{-1}  &  C_{22}^{-1}+C_{22}^{-1}C_{12}^\T\hat C_{11}^{-1}C_{12}C_{22}^{-1}
\end{pmatrix}
\end{equation}
The inverse of a $2\times 2$ correlation matrix reverses the sign of its off-diagonal elements:
\begin{equation}
\begin{pmatrix}
1 & \rho \\
\rho & 1
\end{pmatrix}^{-1}
\propto 
\begin{pmatrix}
-1 & \rho \\
\rho & -1 
\end{pmatrix}
\end{equation}
Therefore, scaling $C^{-1}$ to have 1s on the diagonal yields the partial correlation in each of its $2\times 2$ partitions, resulting in a matrix of partial pairwise correlations.

>>>>

Neural correlations do not come with ready or unambiguous mechanistic interpretations. They can arise from monosynaptic or polysynaptic interactions, common or correlated inputs, oscillations, top-down modulation, and background network fluctuations, and other mechanisms \cite{Perkel:1967, Moore:1970, Shadlen:1998, Salinas:2001, Ostojic:2009, Rosenbaum:2011}. But multineuronal recordings do provide more information than an equivalent number of separately recorded pairs of cells. For example, the eigenvalue decomposition of the covariance matrix expresses shared correlated activity components across the population; common fluctuations of population activity may be accurately represented by only a few eigenvectors that affect all correlation coefficients. On the other hand, a correlation matrix can be specified using the \emph{partial correlations} between pairs of the recorded neurons. The partial correlation coefficient between two neurons reflects their linear association conditioned on the activity of all the other recorded cells.  When a large fraction of the interacting variables are observed, partial correlations more closely reflect the direct causal effects between components of complex systems than correlations \cite{Whittaker:2009}. Partial correlations have been used to describe interactions between genes in functional genomics \cite{Schafer:2005, Peng:2009} and between brain regions in imaging studies \cite{Varoquaux:2012, Ryali:2012}. These opportunities have not yet been explored in neurophysiological studies, as most studies have only considered the distributions of pairwise correlations \cite{Zohary:1994, Bair:2001, Smith:2008, Ecker:2010}.

@@

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.  The usual estimate of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf sample}$ of the empirical sample of observations $x(t),\; t=1,\ldots,n$ as
\begin{equation}\label{eq:sample}
    C_{\sf sample} = \frac 1 \nu \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ denotes degrees of freedom in one neuron's activity within the sample ($\nu=n-1$ if observations are independent). For noise correlations, the true mean $\mu$ and the sample mean $\bar x$ are conditioned on the stimulus. In this study, we estimate the true mean, $\mu$, using the sample mean, $\bar x$, but seek a better estimate of the covariance matrix than the sample correlation matrix, $C_{\sf sample}$.

Given an covariance matrix estimate, the correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2}
\end{equation}
Here $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the identity matrix.

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:2009}. Partial correlations figure prominantly in probablistic \emph{graphical modelling} wherein the joint distribution is explain by sets of two-way interactions \cite{Whittaker:1990}. Under the multivariate Gaussian distribution, zero partial correlations are equivalent to conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1999,Baba:2004}.

Pairwise partial correlations are closely linked to the \emph{precision matrix}, \ie the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1999}. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}
